### 2025-05-13 Chapters/Chapter4.tex
Updated after comments by GNP

\section{Main findings}
As shown in Table \ref{table:setsOfParams}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline as well as a simple regression model. Table \ref{table:results} further demonstrates this improvement. For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline. Table \ref{table:closed_intervals} shows the results of training and model evaluation within closed intervals of the reanalysis wind speed. The best results appear in the 10--25~m/s range. At lower reanlysis wind speeds, percentage errors inflate even when the differences between wind gust and average wind speed remain small and at both at the higher and the lower speeds the base line MAPE is higher.

Although one might expect staility indicators, such as the Richardson number and the Brunt--Väisälä frequency, to be among the strongest predictors of gust variability, they contributed less than other variables. The influence of individual features on model performance is illustrated by Shapley plots (Figure \ref{fig:ShapleySummaryKeflavikurflugvollur} and Appendix \ref{appendix:A}). According the Shapley analysis, stations altitude, mean wind speed and wind direction are among the more important features for predicting the gust factor, and also, to some extent, the Brunt--Väisälä frequency ($N^2$). As explained in Section \ref{sec:shapley} it was infeasible to evaluate the DEM data with Shapley values.

@@

Finally, treating the data as a time series may offer further gains. Here, only instantaneous inputs at forecast time were used, partly because CARRA fields are available every three hours and may not capture rapid stability changes. Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skills.

@@

\section{Possible applications}
Wind gusts at weather stations are displayed on road signs in Iceland, if wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{expected values}. The model could be used to allow such predicitons of wind gusts to be displayed for every section of the road, not just at the stations. In 2023, outside of Reykjavik harbor, a cruise ship nearly \href{https://www.mbl.is/200milur/frettir/2024/04/26/strand_skemmtiferdaskips_kallar_a_erlenda_adstod/}{ran aground} due to an unexpected wind gust. This wind gust of \textasciitilde 26 m/s, pushed the ship into a buoy near the island Viðey. It was only luck that prevented the ship's screws from hitting the buoy. The ship was only 10 meters from the shoals of Viðey, where the water is only 0.4 meters deep. A better prediction of wind gust could have prevented this incidence. Similarly, the model could be used to prevent damage to transportation vehicles, that take on a lot of wind, such as trucks and buses, or to prevent damage to buildings.

>>>>

\section{Main findings}
As shown in Table \ref{table:setsOfParams}, the neural-network model achieves a significant reduction in mean absolute percentage error (MAPE) compared to a constant baseline as well as a simple regression model. Table \ref{table:results} further demonstrates this improvement. For reanalysis wind speeds above 10~m/s, the network reduces error by about 30\% relative to the regression baseline. Table \ref{table:closed_intervals} shows the results of training and model evaluation within closed intervals of the reanalysis wind speed. The best results appear in the 10--25~m/s range. At lower reanalysis wind speeds, percentage errors inflate even when the differences between wind gust and average wind speed remain small and at both at the higher and the lower speeds the base line MAPE is higher.

Although one might expect staility indicators, such as the Richardson number and the Brunt--Väisälä frequency, to be among the strongest predictors of gust variability, they contributed less than other variables. The influence of individual features on model performance is illustrated by Shapley plots (Figures \ref{fig:ShapleySummary}, \ref{fig:ShapleySummaryKeflavikurflugvollur}, \ref{fig:ShapleySummaryAlmannaskarð}, \ref{fig:ShapleySummaryAkrafjall} and figures in Appendix \ref{appendix:A}). According to the Shapley analysis, stations elevation, mean wind speed and wind direction are among the more important features for predicting the gust factor, and also, to some extent, the Brunt--Väisälä frequency ($N^2$). As explained in Section \ref{sec:shapley} it was infeasible to evaluate the DEM data with Shapley values.

@@

Finally, treating the data as a time series may offer further gains. Here, only instantaneous inputs at forecast time were used, partly because CARRA analysis fields are available every three hours and may not capture rapid stability changes. Access to higher-frequency data or the inclusion of temporal history could enable models to exploit evolving atmospheric conditions and further enhance gust-prediction skills.

@@

\section{Possible applications}
Wind gusts at weather stations are displayed on road signs in Iceland, when wind gusts exceed \href{https://www.vegagerdin.is/vegagerdin/starfsemi/frettir/tvo-ny-vedur-og-upplysingaskilti-i-notkun#}{certain values}. The model could be used to allow such predicitons of wind gusts to be displayed for every section of the road, not just at the stations. In 2023, outside of Reykjavik harbor, a cruise ship nearly ran aground due to an unexpected wind gust \parencite{strand_skemmtiferdaskips}. This wind gust of \textasciitilde 26 m/s, pushed the ship into a buoy near the island Viðey. It was only luck that prevented the ship's screws from hitting the buoy. The ship was only 10 meters from the shoals of Viðey, where the water is only 0.4 meters deep. A better prediction of wind gust could have prevented this incidence. Similarly, the model could be used to prevent damage to transportation vehicles, that take on a lot of wind, such as trucks and buses, or to prevent damage to buildings.

### 2025-05-13 MSc.tex
Updated after comments by GNP

\usepackage{biblatex}  % BibLaTeX used for references. 
\usepackage{csquotes} % BibLaTex wants to have context sensitive quotes
\addbibresource{references.bib} %  Name of *.bib file containing references

@@

\section*{Abstract}
Whether it is wind farms, traffic accidents, high-voltage towers, or other things affected by wind, accurately predicting wind loads is crucial for reducing damage, improving operational safety, and boosting productivity. Wind gusts are multi-second spikes in speed, and wind gust factor is defined as the ratio of the highest three-second gust to the ten-minute mean wind speed. It is important to be able to predict both wind and gust factor accurately. Historically, numerical weather prediction models based on physical equations have been used to predict such variables. In recent years, artificial intelligence has advanced rapidly, and AI-based models have begun to rival conventional model performance. This project combines both approaches by using atmospheric model reanalysis output and digital elevation data as inputs to a deep neural network. Selected features from the reanalysis fields, terrain elevation, and derived variables are fed into several network architectures, whose performance is then compared against a simple linear regression baseline. The results are very promising: this hybrid methodology can substantially improve both instantaneous estimation and short-term forecasting of gust factors, particularly in areas of complex terrain.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða aðra hluti sem vindur getur haft áhrif á er spágeta um álag vegna vindsins mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni. Vindhviður eru nokkurra sekúndna aukning á vindi og hviðustuðull mælir hæsta hlutfall milli þriggja sekúndna hviðu og 10 mínútna meðalvinds. Góðar hviðuspár eru mikilvægar. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um veður, þar með talið hviðustuðul. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að meta hviðustuðla. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan sem lýsir landslagi. Djúpt tauganet er búið til með því að velja breytur úr endurgreiningargögnunum og hæðarlíkaninu og afleiddum breytum. Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulega aðhvarfi. Niðurstöðurnar lofa mjög góðu um að þessi aðferðafræði geti bætt bæði samtímamat á hviðustuðli og spár um hann umtalsvert. Þetta á ekki síst við í flóknu landslagi.
%\vfill

@@

\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{Abbreviations}{Abbreviations}
\begin{acronym}[GeoTIFF]
  \acro{API}{Application Programming Interface}
  \acro{ASL}{Above Sea Level}
  \acro{AWS}{Automatic Weather Stations}
  \acro{CARRA}{Copernicus Artic Regional ReAnalysis dataset}
  \acro{CNN}{Convolutional Neural Networks}
  \acro{DEM}{Digital Elevaiton Model}
  \acro{ELI5}{Explain Like I am 5}
  \acro{ECMWF}{European Centre for Medium-Range Weather Forecast}
  \acro{GCM}{General Circulation Model}
  \acro{GeoTIFF}{Georeferenced TIFF}
  \acro{GPU}{Graphical Processing Unit}
  \acro{HRES}{High Resolution forecas}
  \acro{IMO}{Icelandic Meteoroligcal Office}
  \acro{IRCA}{Icelandic Road and Coastal Administration}
  \acro{JNWPU}{Joint Numerical Weather Prediction Unit}
  \acro{LWM}{Large AI Weather forecast Model}
  \acro{MAE}{Mean Asbolute Error}
  \acro{MAPE}{Mean Absolute Percentage Error}
  \acro{NN}{Neural Network}
  \acro{NWP}{Numerical Weather Prediction}
  \acro{SENS}{School of Engineering and Natural Sciences}
  \acro{TIFF}{Tag Image File Format}
  \acro{UoI}{University of Iceland}
\end{acronym}

>>>>

\usepackage[style=apa, backend=biber]{biblatex}  % BibLaTeX used for references. 
\usepackage{csquotes} % BibLaTex wants to have context sensitive quotes
\addbibresource{references.bib} %  Name of *.bib file containing references

@@

\section*{Abstract}
Whether it is wind farms, traffic accidents, high-voltage towers, or other things affected by wind, accurately predicting wind loads is crucial for reducing damage, improving operational safety, and boosting productivity. Wind gusts are spikes in wind speed, and wind gust factor is defined as the ratio of the highest gust to the mean wind speed over a given time period. It is important to be able to predict both wind and gust factor accurately. Historically, numerical weather prediction models based on physical equations have been used to predict such variables. In recent years, artificial intelligence (AI) has advanced rapidly, and AI-based models have begun to rival conventional model performance. This project combines both approaches by using atmospheric model reanalysis output and digital elevation data as inputs to a deep neural network. Selected features from the reanalysis fields, terrain elevation, and derived variables are fed into several network architectures, whose performance is then compared against a simple linear regression baseline. The results are very promising: this hybrid methodology can substantially improve both instantaneous estimation and short-term forecasting of gust factors, particularly in areas of complex terrain.
\vfill \vspace*{1cm}

\section*{Útdráttur}
Hvort sem um er að ræða vindmyllugarða, umferðarslys, háspennumöstur, eða annað sem vindur getur haft áhrif á er spágeta um álag vegna vinds mikilvæg, til að minnka eða koma í veg fyrir tjón, auka rekstraröryggi, eða bæta framleiðni. Vindhviður eru toppar í vindhraða og hviðustuðull er hlutfall milli hæstu vindhviðu og meðalvinds yfir gefið tímabil. Góðar hviðuspár eru mikilvægar. Sögulega hefur verið notast við hefðbundin spálíkön sem reiða sig á töluleg eðlisfræðileg líkön til að spá fyrir um veður, þar með talið hviðustuðul. Á síðustu árum hefur gervigreind þróast mikið og líkön verið þjálfuð til þess að spá fyrir um veður. Svo miklar framfarir hafa orðið að sum líkön geta keppt við hefðbundin veðurlíkön. Markmið þessa verkefnis er reyna að nýta bæði hefðbundin veðurlíkön og gervigreind til þess að meta hviðustuðla. Þetta er gert með því að nota endurgreiningargögn úr hefðbundnu veðurlíkani sem grunngögn í gervigreindarlíkan. Einnig eru notað hæðarlíkan sem lýsir landslagi. Djúpt tauganet er búið til með því að velja breytur úr endurgreiningargögnunum og hæðarlíkaninu og afleiddum breytum. Nokkur líkön af þessu tagi eru svo þjálfuð og niðurstöður bornar saman við einfalt grunnlíkan sem fengið er með línulegu aðhvarfi. Niðurstöðurnar lofa mjög góðu um að þessi aðferðafræði geti bætt bæði samtímamat á hviðustuðli og spár um hann umtalsvert. Þetta á ekki síst við í flóknu landslagi.
%\vfill

@@

\chapter*{Abbreviations}
\addcontentsline{toc}{chapter}{Abbreviations}
\markboth{Abbreviations}{Abbreviations}
\begin{acronym}[GeoTIFF]
  \acro{API}{Application Programming Interface}
  \acro{ASL}{Above Sea Level}
  \acro{AWS}{Automatic Weather Stations}
  \acro{CARRA}{Copernicus Artic Regional ReAnalysis}
  \acro{CNN}{Convolutional Neural Networks}
  \acro{DEM}{Digital Elevaiton Model}
  \acro{ELI5}{Explain Like I am 5}
  \acro{ECMWF}{European Centre for Medium-Range Weather Forecast}
  \acro{GCM}{General Circulation Model}
  \acro{GeoTIFF}{Georeferenced TIFF}
  \acro{GPU}{Graphical Processing Unit}
  \acro{GRIB}{GRIdded Binary}
  \acro{HRES}{High Resolution forecast}
  \acro{IMO}{Icelandic Meteorological Office}
  \acro{IRCA}{Icelandic Road and Coastal Administration}
  \acro{JNWPU}{Joint Numerical Weather Prediction Unit}
  \acro{LWM}{Large AI Weather forecast Model}
  \acro{MAE}{Mean Asbolute Error}
  \acro{MAPE}{Mean Absolute Percentage Error}
  \acro{NN}{Neural Network}
  \acro{NWP}{Numerical Weather Prediction}
  \acro{SENS}{School of Engineering and Natural Sciences}
  \acro{TIFF}{Tag Image File Format}
  \acro{UoI}{University of Iceland}
\end{acronym}
